---
title: "Homework 2"
output: github_document
date: "2025-09-30"
---

Loading the library first: 

```{r}
library(tidyverse)
library(lubridate)
```

Getting the exact path:
```{r}
getwd()
```

Getting the data frames up on the R:
```{r}
pols_month = 
  read_csv("/Users/karolinawiesiolek/Desktop/Data_Science/Homeworks/Homework2/data/pols-month.csv") 

unemployment = 
  read_csv("/Users/karolinawiesiolek/Desktop/Data_Science/Homeworks/Homework2/data/unemployment.csv") 

snp = 
  read_csv("/Users/karolinawiesiolek/Desktop/Data_Science/Homeworks/Homework2/data/snp.csv") 

```

## Working on pols_month -- first part of Problem 1.

Step 1. Separate by year/month/day.
Step 2. Replace month number with a month name.
Step 3. Create 'president' variable taking values gop and dem. First, I used a second mutate to add new variable but that didn't work so i merged 'mutate'.
Step 4. Using `select` removed columns. 

```{r}
pols_month =
  read_csv("/Users/karolinawiesiolek/Desktop/Data_Science/Homeworks/Homework2/data/pols-month.csv")
  pols_month_clean = pols_month %>% 
  separate(mon, into = c("year", "month", "day"), sep = "-", convert = TRUE) %>% 
  mutate(
    month = month.name[month],
    president = case_when(
    prez_gop == 1 ~ "gop",
    prez_dem == 1 ~ "dem"
  )
) %>% 

select(-prez_dem, -prez_gop, -day) 
 
```



## 2nd part of Problem 1 -- snp.csv 
Step 1. Arrange according to year/month - these are first columns. 

```{r}
snp =
  read_csv("/Users/karolinawiesiolek/Desktop/Data_Science/Homeworks/Homework2/data/snp.csv")
  snp_clean = snp %>% 
  separate(date, into = c("month", "day", "year"), sep = "-", convert = TRUE) %>% 
  mutate(month = month.name[month]) %>% 
  select(year, month, everything()) %>%
  arrange(year, match(month, month.name))

 ```
